---
title: "Dumbbell workout efficiency analysis"
author: "Bertrand Dolimier"
date: "12/14/2016"
output:
  html_document: default
  pdf_document: default
---

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. 
  
One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.  

Approach: We will split the training dataset into a "real" training and validation datasets. Will explore liner regression as a method to resolve the testing dataset (even if all indicators lead us to believe this is a classification problem). If LR is not conclusive we will still use result to select the fields correlated with the "classe" field to apply to the subsequent methods. We will then try Tree, Random Forest, Bagging and Boosting, we will keep the method with best accuracy results on the validation dataset and apply that method on the testing dataset for final prediction.


# Libraries / Loading raw data / Inititiatization 
We will split the training dataset into a training of 10% and validation 90%
```{r loading_libs, include=TRUE, warning=FALSE, message=FALSE}
library(ggplot2)
library(lattice)
library( caret )
library(kernlab)
library(corrplot)
library(dplyr)
library(rpart)
library(party)
library(mboost)

CSVtraining = read.csv("../pml-training.csv" , na.strings=c("NA","#DIV/0!","") )
testing = read.csv("../pml-testing.csv" , na.strings=c("NA","#DIV/0!","") )
inTrain <- createDataPartition(y=CSVtraining$classe,
                              p=0.1, list=FALSE)
training <- CSVtraining[inTrain,]
validation <- CSVtraining[-inTrain,]

#dim(CSVtraining)
#dim(training)
#dim(validation)
#dim(testing)

```

# Linear Regression exploration  #
Select training numeric fields not null and with minimum correlation with the classe column
  Display plot to confirm fields selected are correlated

We will first remove all fields with no significant values in the test dataset, then explore the correlation betwen the remaining fields and the classe field and keep on a selected fields. Listed in this code chunk.

```{r corrplot, echo=FALSE}
useColumns<-as.logical(!is.na(training[1,]))
trainData<-training[,useColumns]
nzv<- nearZeroVar(trainData,saveMetrics = TRUE)
head(nzv)

df <- cbind(
  training["magnet_belt_y" ]	
, training["magnet_belt_z" ]	
, training["roll_arm" ]	
, training["pitch_arm" ]	
, training["yaw_arm" ]	
, training["total_accel_arm" ]
, training["accel_arm_x" ]	
, training["accel_arm_y" ]	
, training["accel_arm_z" ]	
, training["magnet_arm_x" ]	
, training["magnet_arm_y" ]	
, training["magnet_arm_z" ]
, training["pitch_forearm" ]
, training["yaw_forearm" ]
, training["total_accel_forearm" ]
, training["gyros_forearm_x" ]
, training["magnet_forearm_x" ]	
, training["magnet_forearm_y" ]	
 )

# Make the factor classe column a numeric for linear analysis
fac         <- as.factor( training$classe )
df$classe   <- as.numeric(fac)

# Substitute zero to null
df[is.na( df )] <- 0
df[is.null( df )] <- 0

# Correlation plot
mcor <- cor( df )
corrplot.mixed( mcor , tl.col="black", tl.srt=25)
```

# Linear Regression plot
The plot below clearly shows a correlation between the 15 combined predictors but also its predictive limitation as fit overlap is large accross all 5 classes.
  As suspected, as classification model is far better appropriate.
  
```{r ,echo=FALSE}
df <- cbind(
  training["magnet_belt_y" ]	
, training["magnet_belt_z" ]	
, training["roll_arm" ]	
, training["pitch_arm" ]	
, training["yaw_arm" ]	
, training["total_accel_arm" ]
, training["accel_arm_x" ]	
, training["accel_arm_y" ]	
, training["accel_arm_z" ]	
, training["magnet_arm_x" ]	
, training["magnet_arm_y" ]	
, training["magnet_arm_z" ]
, training["pitch_forearm" ]
, training["yaw_forearm" ]
, training["total_accel_forearm" ]
, training["gyros_forearm_x" ]
, training["magnet_forearm_x" ]	
, training["magnet_forearm_y" ]	
 )
fac         <- as.factor( training$classe )
df$classe   <- as.numeric(fac)

# Plot fitted value with classe
fit <- lm( classe ~ ., df )
plot( df$classe, fit$fitted,pch=19,col="blue",xlab="Class",ylab="Fit...")

#boxplot(classnum ~ fit$fitted.values ,data=df, main="Dumb bell effectiveness Analysis", xlab="Average Pitch Belt", ylab="Classification")
# Reset classe to its factor values
df$classe <- training$classe

```

We selected 15 fields from the training dataset base on their correlation showed on the plot above.
  That subset will be used as predictors for the methods below.
  
# Tree #  
First method explore, the "tree" method aka "rpart" in the caret package.
We find the accuracy between on the validation dataset to be 40% which clearly not satisfactory.
```{r arbre, echo=FALSE}
modFit <- train( classe ~  (
  magnet_belt_y	
+ magnet_belt_z	
+ pitch_arm	
+ total_accel_arm
+ accel_arm_x	
+ accel_arm_y	
+ accel_arm_z	
+ magnet_arm_x	
+ magnet_arm_y	
+ magnet_arm_z
+ pitch_forearm
+ yaw_forearm
+ total_accel_forearm
+ magnet_forearm_x	
+ magnet_forearm_y )
 ,data=df ,method="rpart")
modFit

# Predict on validation dataset
validation[is.na( validation )] <- 0
validation[is.null( validation )] <- 0
pred <- predict(modFit, validation)
table(pred==validation$classe,  validation$classe)
table(pred,  validation$classe)

#qplot( pred, validation$classe,colour=classe,data=validation)
print(modFit$finalModel)
```

# Bagging #
 Plot shows classe "A" (exercice done correctly) has the smallest range of fitted values, however classification of "bad" workout is not resolved.
```{r bag, echo=FALSE}
predictors  = data.frame(classe=df$classe)
belts_measures = (
  df$magnet_belt_y	
+ df$magnet_belt_z	
+ df$pitch_arm	
+ df$total_accel_arm
+ df$accel_arm_x	
+ df$accel_arm_y	
+ df$accel_arm_z	
+ df$magnet_arm_x	
+ df$magnet_arm_y	
+ df$magnet_arm_z
+ df$pitch_forearm
+ df$yaw_forearm
+ df$total_accel_forearm
+ df$magnet_forearm_x	
+ df$magnet_forearm_y )

treebag <- bag(predictors, belts_measures, B = 10,
                bagControl = bagControl(fit = ctreeBag$fit,
                                        predict = ctreeBag$pred,
                                        aggregate = ctreeBag$aggregate))

plot(df$classe,belts_measures ,col='lightgrey',pch=19)

```


# Random Forest #
Best model so far, has best accuracy.
```{r, echo=FALSE}
modFit <- train( classe ~ (
  magnet_belt_y	
+ magnet_belt_z	
+ pitch_arm	
+ total_accel_arm
+ accel_arm_x	
+ accel_arm_y	
+ accel_arm_z	
+ magnet_arm_x	
+ magnet_arm_y	
+ magnet_arm_z
+ pitch_forearm
+ yaw_forearm
+ total_accel_forearm
+ magnet_forearm_x	
+ magnet_forearm_y )
,data=df ,method="rf",prox=TRUE)
modFit

# Predict on validation dataset
pred <- predict(modFit, validation)
table(pred==validation$classe,  validation$classe)
table(pred,  validation$classe)

SVmodFit <- modFit
```

# Boosting #
Accuracy of this method is below 50%, we shall not select this model.
```{r boosting, echo=FALSE}
modFit <- train( classe ~ 
                   (
  magnet_belt_y	
+ magnet_belt_z	
+ pitch_arm	
+ total_accel_arm
+ accel_arm_x	
+ accel_arm_y	
+ accel_arm_z	
+ magnet_arm_x	
+ magnet_arm_y	
+ magnet_arm_z
+ pitch_forearm
+ yaw_forearm
+ total_accel_forearm
+ magnet_forearm_x	
+ magnet_forearm_y )
, method="gbm",data=training,verbose=FALSE)

print(modFit)
# Predict on validation dataset
pred <- predict(modFit, validation)
table(pred==validation$classe,  validation$classe)
table(pred,  validation$classe)
```

# Result #
We selected Random Forest as best model.
Prediction results on the testing dataset are:
```{r}
# Predict on testing set
testing[is.na( testing )] <- 0
testing[is.null( testing )] <- 0
pred <- predict(SVmodFit, testing)
pred
```